---
title: "Problem_Set_3"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: no
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This problem set will be due Thursday March 2rd, as mentioned in class please submit this homework as an .R file

#### For this assignment and any future homework the fundamental aspect of an acceptable answer is the analysis and conclusions of your analysis. This means that for every answer please write a complete and concise conclusion (not a long one, just a clear one)

#### In this assignment we are going to review the concepts from class this week. However, concepts build on each other for every test you run you need to make sure that the data meets the appropiate criteria to be able to run the test ####Important: Use comments throughout your code to:

#### Indicate which question you are answering

#### Example: #Question 1A, #Question 1B, etc

#### Annotate your work

#### Examples: #Calculate the mean for 'height'

##### P(being on time) = 1/6

##### Apply binomial distribution

#### If you do not provide annotation, we will not be able to give you full credit for solving the problems.

#### If you have any questions about what is being asked or what you need to do in order to solve the problems, please reach out to a TA or instructor as soon as possible. If it is necessary to provide additional information or corrections, updated information will be posted on the HW3 Assignment page in Canvas.

### Dataset:

### For questions A and B, use the dataset for "Risk Factors Associated with Low Infant Birth Weight." You should use R to examine the dataset and answer the following questions:

#### Additional information about the dataset:

#### Source: Hosmer, D.W. and Lemeshow, S. (1989) Applied Logistic Regression New York-Wiley

#### This data frame contains the following columns

#### low: indicator of birth weight less than 2.5 kg

#### age: mother's age in years

#### lwt: mother's weight in pounds at last menstrual period

#### race: mother's race ('1' = white, '2' = black, '3' = other)

#### smoke: smoking status during pregnancy. 1 = yes, 0 = no

#### ptl: number of previous premature labours.

#### ht: history of hypertension.

#### ui: presence of uterine irritability.

#### ftv: number of physician visits during the first trimester.

#### bwt: birth weight in grams

```{r dataset}

#birthwt <- read.table("birthwt.txt", sep="",header = T)

#apply(birthwt,2,table) ##Information regarding the variables
```

## 20 pts A. It is thought that smoking affects fetus growth and development and we want to test whether smoking during pregnancy reduces fetal weight. You want to test this idea.

#### 2 pts (i) What kind of statistical test should you use? Is this a parametric or non-parametric test, and what are the specific assumptions about data that underlie this test?

#### 1 pts (ii) What is your null hypothesis?

#### 1 pts (iii) What is your alternative hypothesis?

#### 5 pts (iv) Run the test.

#### 1 pts (v) What is your conclusion?

```{r a}
### A. It is thought that smoking affects fetus growth and development and we want to test whether smoking during pregnancy reduces fetal weight. You want to test this idea.
```

## 20 pts B. You want to compare the birth weights between individuals whose mothers identified as different races in this study.

#### 2 pts (i) What kind of statistical test should you use? Is this a parametric or non-parametric test, and what are the specific assumptions about data that underlie this test?

#### 1 pts (ii) What is your null hypothesis?

#### 1 pts (iii) What is your alternative hypothesis?

#### 5 pts (iv) Run the test/tests.

#### 1 pts (v) What is your conclusion, with respect to which pairs of specific groups have significant differences in birth weight?

```{r b}
### B. You want to compare the birth weights between individuals whose mothers identified as different races in this study.
```

## 15 pts C. A study was conducted to investigate the effect of physical training on the serum cholesterol level.

### Thirty subjects participated in the study. Prior to training, blood samples were taken to determine the cholesterol level of each subject. Then the subjects were put through a training program that centered on daily running and jogging. At the end of the training period, blood samples were taken again and a second reading on the serum cholesterol level was obtained.

#### 3 pts (i) What kind of statistical test should you use? Is this a parametric or non-parametric test, and what are the specific assumptions about data that underlie this test?

#### 1 pts (ii) What is your null hypothesis?

#### 1 pts (iii) What is your alternative hypothesis?

#### 5 pts (iv) Run the test/tests.

#### 5 pts (v) What is your conclusion, Is there an effect of the training on cholesterol levels?

```{r c}
### Is there an effect of the training on cholesterol levels?
### The data is located in this working directory with the name training.csv

##Load data
#training = read.csv("training.csv")
```

## 15 pts. D. A group of 24 patients with chronic pain was randomized into three groups receiving different treatments.

### The pain level (on a scale from 1 to 10) was measured after 1 week of treatment. Test the hypothesis that there is no difference in pain level among the three treatments test. Use a significance level of 0.05.

#### 1.Treatment A: 7, 6, 8, 9, 5, 6, 7, 8

#### 2.Treatment B: 4, 5, 6, 4, 7, 6, 5, 4

#### 3.Treatment C: 2, 1, 3, 4, 3, 2, 1, 2

#### 3 pts (i) What kind of statistical test should you use? Is this a parametric or non-parametric test, and what are the specific assumptions about data that underlie this test?

#### 1 pts (ii) What is your null hypothesis?

#### 1 pts (iii) What is your alternative hypothesis?

#### 5 pts (iv) Run the test/tests.

#### 5 pts (v) What is your conclusion, with respect to which pairs of specific groups have significant differences in pain level (hint: pairwise.wilcox.test function in R)?

```{r bwt}
a = c(7, 6, 8, 9, 5, 6, 7, 8 )
b = c(4, 5, 6, 4, 7, 6, 5, 4 )
c = c(2, 1, 3, 4, 3, 2, 1, 2)

df = data.frame(vals = c(a,b,c),treat = c(rep("a",8),rep("b",8),rep("c",8)))

```

## The following questions do not require R.

## 10 pts E. Distinguish between parametric vs non-parametric tests and explain the tradeoffs in choosing one vs the other for hypothesis testing.

## 20 pts. F. A survey was conducted to determine the average income of households in a certain city.

#### A random sample of 500 households was selected, and the mean income was found to be \$60,000, with a standard deviation of \$10,000.

#### 7 pts (i) Calculate the standard error of the sample mean.

#### 7 pts (ii) What is the 95% confidence interval for the population mean income?

#### 6 pts (iii) How is the interpretation of the value obtained for the standard error different from the one obtained for the standard deviation?

```{r e}
###SE = 
###CI =
```
